Predicting English Premier League Rankings in Relation to Total Team Spending in the 2023-2024 Season

### 2024-04-03 paper/paper.qmd
Committing all work done so far to paper.gmd and changes to analysis_data

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

merged_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
show_col_types = FALSE
```

@@

# Data {#sec-data}
```{r plot-chunk, fig.cap="Correlation matrix of numeric variables", fig.width=8, fig.height=7, echo=FALSE}
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Assuming 'lg_rank' is your league rank variable
merged_data$transformed_rank <- max(merged_data$lg_rank) + 1 - merged_data$lg_rank
# Calculate the correlation matrix for numeric columns
cor_matrix <- cor(merged_data |> select(where(is.numeric)), use = "complete.obs")

corrplot(cor_matrix, method = "circle", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45, tl.cex = 0.7)

# Manually add title
mtext("Correlation Matrix for Toronto FC's Performance Features", side=3, line=3, cex=1.5)
```

@@

@@

```{r}
#| label: tbl-lg-rank-correlation
#| tbl-cap: Correlation of Variables with lg_rank
#| echo: false
#| warning: false

# Convert the correlation matrix to a long-format dataframe, focusing on lg_rank
cor_df_long_lg_rank <- cor_matrix |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  pivot_longer(cols = -Variable, names_to = "Related_Variable", values_to = "Correlation") |>
  filter(Variable == "lg_rank" ) |>
  distinct()

# Use knitr::kable to display the table, focusing on correlations with lg_rank
knitr::kable(cor_df_long_lg_rank, padding = 5, format = "latex", booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = c("striped", "scale_down"))
```

>>>>

---
title: "Predicting English Premier League Rankings in Relation to Total Team Spending in the 2023-2024 Season"
subtitle: "My subtitle if needed"
author: 
  - Isabella MacLean
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
show_col_types = FALSE

matchday_attendance <- read_csv(here::here("data/raw_data/average_home_matchday_attendance_2023-2024.csv"))
show_col_types = FALSE

market_value <- read_csv(here::here("data/raw_data/market_value_2023-2024.csv"))
show_col_types = FALSE

team_rank <- read_csv(here::here("data/raw_data/team_rank_2023-2024.csv"))
show_col_types = FALSE

total_payroll <- read_csv(here::here("data/raw_data/total_payroll_2023-2024.csv"))
show_col_types = FALSE

total_transfer_spending <- read_csv(here::here("data/raw_data/total_transfer_spending_2023-2024.csv"))
show_col_types = FALSE
```

@@

# Data {#sec-data}

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-pts_vs_average_home_matchday_attendance
#| fig-cap: Figure 1 presents a scatter plot illustrating the relationship between total league points and average matchday attendance during the 2023-2024 season. Each point represents a team, plotted with their total points on the x-axis and their average home attendance on the y-axis. The trend line, added using linear regression analysis, suggests a correlation between team performance and matchday turnout. The correlation coefficient of 0.694060419947882 indicates a positive association with moderate correlation, where teams with higher league points tend to have greater average attendance.

# Calculate correlation coefficient for league position and average attendance
cor_attendance <- cor(analysis_data$pts, analysis_data$average_home_matchday_attendance, use = "complete.obs")

# Create a scatter plot for league position and average attendance
ggplot(analysis_data, aes(x = pts, y = average_home_matchday_attendance)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Figure 1: Average Matchday Attendance in Relation 
       to League Position for the 2023-2024 Season",
       subtitle = paste("Correlation coefficient:", cor_attendance),
       x = "Total League Points",
       y = "Average Attendance")
```

```{r}
#| warning: false
#| echo: false
#| label: fig-pts_vs_total_wage_bill
#| fig-cap: Figure 2 depicts the association between total league points and the total payroll for teams in the 2023-2024 season. Each point on the plot corresponds to a team, positioned according to their accumulated points for the season along the x-axis and their respective payroll on the y-axis. The linear regression trend line suggests a correlation between a team’s financial expenditure on payroll and their success in the league. A correlation coefficient of 0.770829037567151 reflects a positive relationship, implying that teams with higher payrolls may be more likely to accumulate greater league points.

# Calculate correlation coefficient for league position and average total wages
cor_wages <- cor(analysis_data$pts, analysis_data$total_wage_bill, use = "complete.obs")

# Create a scatter plot for league position and average total wages
ggplot(analysis_data, aes(x = pts, y = total_wage_bill)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Figure 2: Total Payroll in Relation to League 
       Points for the 2023-2024 Season",
       subtitle = paste("Correlation coefficient:", cor_wages),
       x = "Total League Points",
       y = "Total Payroll")
```

```{r}
#| warning: false
#| echo: false
#| label: fig-pts_vs_transfer_fees
#| fig-cap: Figure 3 illustrates the relationship between teams total transfer spending and their corresponding league points for the 2023-2024 season. The scatter plot shows each team as a distinct point, positioned by their total league points on the x-axis and their transfer spending on the y-axis. The trend line, calculated with linear regression, indicates a moderate positive correlation, with a correlation coefficient of approximately 0.417985956450557. This suggests that higher league points tend to coincide with increased transfer spending, although the relationship is not strongly linear.

# Calculate correlation coefficient for league position and total transfer fees
cor_transfers <- cor(analysis_data$pts, analysis_data$transfer_fees, use = "complete.obs")

# Create a scatter plot for league position and average total wages
ggplot(analysis_data, aes(x = pts, y = transfer_fees)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Figure 3: Average Transfer Spending in Relation to 
       League Points for the 2023-2024 Season",
       subtitle = paste("Correlation coefficient:", cor_transfers),
       x = "Total League Points",
       y = "Total Transfer Fees")
```
```{r}
#| warning: false
#| echo: false
#| label: fig-pts_vs_market_value
#| fig-cap: Figure 4 demonstrates the correlation between the current market value of teams and their accumulated league points for the 2023-2024 season. The scatter plot places each team by the total points they have secured on the x-axis against their market value on the y-axis. A pronounced upward trend represented by the linear regression line, with a correlation coefficient of 0.849962012423038, indicates a strong positive relationship. This suggests that teams with higher market values tend to amass more league points, underlining a potential link between financial strength and on-field success.

# Calculate correlation coefficient for league position and market value
cor_market <- cor(analysis_data$pts, analysis_data$market_value, use = "complete.obs")

# Create a scatter plot for league position and market value
ggplot(analysis_data, aes(x = pts, y = market_value)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Figure 4: Current Market Value in Relation 
       to League Points for the 2023-2024 Season",
       subtitle = paste("Correlation coefficient:", cor_market),
       x = "Total League Points",
       y = "Total Market Value")
```

@@
